Fusing Filters with Integer Linear Programming

### 2014-05-16 2014betterfusionforfilters/section/05-Benchmarks.tex
Copy edits to S5.1: normalize

\begin{tabular}{lrrrrrrrr}
                &   \multicolumn{2}{c}{Unfused}         & \multicolumn{2}{c}{Stream}
                & \multicolumn{2}{c}{Megiddo} &\multicolumn{2}{c}{\textbf{Ours}} \\
                & Time & Loops   & Time & Loops      & Time & Loops & Time & Loops   \\
\hline
Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
Closest pair    & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
\end{tabular}

@@

\subsection{Normalize2}
To demonstrate the ILP formulation, the clustering of @normalize2@ is derived as follows.
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (>   0)  xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}

To make the program shorter and easier to work with, we use the optimised version.
First, we must calculate $possible$ -- that is, those nodes which have no fusion-preventing path between them.
Any elements in the following sets have no fusion-preventing paths.
\[ \{ \{sum1, gts, sum2\}
 , \{sum1, ys2\}
 , \{gts, sum2, ys1\}
 , \{ys1, ys2\} \} \]

Note that, in the objective function, the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $ 25 \cdot x_{ys1, ys2} + 5 \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
Subject to \\
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ \\
\\
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ \\
\\
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$ \\
\\
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$ \\
\\
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}

One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts}, x_{sum1, sum1}, x_{sum1, sum2}, x_{gts, sum2}, x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$ \\
\\
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
\\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
This minimal solution is not unique, however in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering.

Looking at just the non-zero variables in the objective function, the value is
$25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$.
It is worth noting that, for example, the objective could be reduced by fusing $x_{sum1,ys2}$ together.
However, this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, the constraint requires that $\pi_{sum1} = \pi_{sum2}$, and another constraint requires $\pi_{sum2} < \pi_{ys2}$.
These constraints may not all hold, so a clustering that fused sum1 and ys2 together would not allow sum1 and sum2 to be fused together.

We can also calculate the objective function for the clustering used by stream fusion, to contrast and show that the constraints are valid.
To do this, the values of $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}
If we add these extra equality constraints to the integer linear program and solve it, one possible solution is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
It is interesting to note that two nodes having equal $\pi_i$ values does not imply that the two nodes are fused together.
Conversely, however, different $\pi_i$ values do imply that two nodes are not fused together.

The objective function's value is
$25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.

\subsection{Closest Points}
The closest points benchmark is a divide-and-conquer algorithm that, given an array of points, finds the closest pair of points.
A midpoint is found, and the points are filtered into those above the midpoint, and those below.
The closest points in the filtered arrays are then recursively computed, and the two results merged.
As the filters are operated on recursively, there are no opportunities for fusing the filters, making our clustering the same as Megiddo's.
Our clustering is able to generate both filtered arrays in a single loop, unlike stream fusion, which requires a separate loop for each.

\subsection{QuadTree}
The QuadTree benchmark recursively computes a 2-dimensional space partitioning tree of an array of points.
At each step, the array of points is filtered into four boxes.
As with closest points, there are no opportunities for fusing the filters, and our clustering is the same as Megiddo's.
Our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.

\subsection{QuickHull}
QuickHull is another program that is worth mentioning, as it highlights some fusion opportunities with filters.
The core of the QuickHull algorithm is shown below: given a line and an array of points, the points are filtered to those above the line, and the farthest point above the line is also found.

>>>>

\begin{tabular}{lrrrrrrrr}
                & \multicolumn{2}{c}{Unfused}         & \multicolumn{2}{c}{Stream}
                & \multicolumn{2}{c}{Megiddo} &\multicolumn{2}{c}{\textbf{Ours}} \\
                & Time & Loops   & Time & Loops      & Time & Loops & Time & Loops   \\
\hline
Normalize2      & 1.88s & 5      & 1.64s & 4          & 1.82s & 3  & \textbf{1.59s} & \textbf{2}\\
Closest points  & 3.83s & 6      & 3.33s & 5          & 2.92s & 3  & \textbf{2.92s} & \textbf{3}\\
QuadTree        & 5.22s & 8      & 5.22s & 8          & 4.72s & 2  & \textbf{4.72s} & \textbf{2}\\
\end{tabular}

@@

% -----------------------------------------------------------------------------
\subsection{Normalize2}
To demonstrate the ILP formulation we will use the @normalize2@ example from \S\ref{s:Introduction}, repeated here:
\begin{code}
  normalize2 :: Array Int -> Array Int
  normalize2 xs
   = let sum1 = fold   (+)  0   xs
         gts  = filter (>   0)  xs
         sum2 = fold   (+)  0   gts
         ys1  = map    (/ sum1) xs
         ys2  = map    (/ sum2) xs
     in (ys1, ys2)
\end{code}
We use the final ILP formulation from \S\ref{s:OptimisedConstraints}. First, we calculate $possible$ -- that is, the nodes which have no fusion-preventing path between them.
\[ \{ \{sum1, gts, sum2\}
 , \{sum1, ys2\}
 , \{gts, sum2, ys1\}
 , \{ys1, ys2\} \} \]

The complete ILP program is shown below. Note that in the objective function the weights for $x_{sum1, sum2}$ and $x_{sum2, ys1}$ are both only 1, because they do not share any input arrays.

\begin{tabbing}
MMMMM   \= MMMMMMM \= M \= MMMMMMM \= M \= MMMMMMM \= \kill
Minimise   \> $25 \cdot x_{sum1, gts} + 1  \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} +$ \\
           \> $25 \cdot x_{gts, sum2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} +$ \\
           \> $25 \cdot x_{ys1, ys2}  + 5  \cdot c_{gts} + 5 \cdot c_{ys1} + 5 \cdot c_{ys2} $\\
Subject to \\
    \> $-5 \cdot x_{sum1, gts}$  \> $\le$ \> $\pi_{gts} - \pi_{sum1}$  \> $\le$ \> $5 \cdot x_{sum1, gts}$  \\
    \> $-5 \cdot x_{sum1, sum2}$ \> $\le$ \> $\pi_{sum2} - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, sum2}$ \\
    \> $-5 \cdot x_{sum1, ys2 }$ \> $\le$ \> $\pi_{ys2 } - \pi_{sum1}$ \> $\le$ \> $5 \cdot x_{sum1, ys2 }$ \\
    \> $-5 \cdot x_{gts,  ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, ys1  }$ \\
    \> $-5 \cdot x_{sum2, ys1 }$ \> $\le$ \> $\pi_{ys1 } - \pi_{sum2}$ \> $\le$ \> $5 \cdot x_{sum2, ys1 }$ \\
    \> $-5 \cdot x_{ys1, ys2  }$ \> $\le$ \> $\pi_{ys2 } - \pi_{ys1 }$ \> $\le$ \> $5 \cdot x_{ys1, ys2  }$ 
\\[1ex]
    \> $   x_{gts, sum2 }$ \> $\le$ \> $\pi_{sum2} - \pi_{gts }$ \> $\le$ \> $5 \cdot x_{gts, sum2 }$ 
\\[1ex]
    \>                     \>       \> $\pi_{sum1} < \pi_{ys1}$ \\
    \>                     \>       \> $\pi_{sum2} < \pi_{ys2}$
\\[1ex]
    \> $ x_{gts,sum2} $    \> $\le$ \> $c_{gts}$
\\[1ex]
    \> $x_{gts, sum2}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1,sum1}$     \> $\le$ \> $x_{sum1, sum2}$ \\
    \> $x_{sum1, gts}$     \> $\le$ \> $x_{sum1, sum2}$
\end{tabbing}

\noindent
One minimal solution to this is:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{sum1, gts},~ x_{sum1, sum1},~ x_{sum1, sum2},~ x_{gts, sum2},~ x_{ys1,  ys2}$
    \> $=$ \> $0$ \\
$x_{sum1, ys2},~ x_{gts, ys1},~   x_{sum2, ys1}$
    \> $=$ \> $1$ 
\\[1ex]
$\pi_{sum1},~ \pi_{gts },~ \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 },~ \pi_{ys2 }$
    \> $=$ \> $1$ 
\\[1ex]
$c_{gts},~ c_{ys1},~ c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
This minimal solution is not unique, though in this case the only other minimal solutions use different $\pi$ values, and denote the same clustering. Looking at just the non-zero variables in the objective function, the value is $25 \cdot x_{sum1,ys2} + 25 \cdot x_{gts,ys1} + 1 \cdot x_{sum2, ys1} = 51$. For illustrative purposes, note that objective function could be reduced by setting $x_{sum1,ys2} = 0$ (fusing $sum1$ and $ys1$), but this conflicts with other constraints. Since $x_{sum1, sum2} = 0$, we require $\pi_{sum1} = \pi_{sum2}$, but also \mbox{$\pi_{sum2} < \pi_{ys2}$}. These constraints cannot be satisfied, so a clustering that fused $sum1$ and $ys2$ would not also permit $sum1$ and $sum2$ to be fused.

We will now compare the clustering produced by our system, with the one implied by stream fusion. As described in \cite{lippmeier2013flow}, stream fusion cannot fuse a produced array into multiple consumers, or fuse operators that are not in a producer-consumer relationship. The corresponding values of the $x_{ij}$ variables are:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$x_{gts, sum2}$
    \> $=$ \> $0$ \\
$x_{sum1, gts}, x_{sum1, sum2}, x_{ys1,  ys2}, x_{sum1, ys2}, x_{gts, ys1 }, x_{sum2, ys1}$
    \> $=$ \> $1$
\end{tabbing}
We can force this clustering to be applied in our integer linear program by adding the above equations as new constraints. Solving the resulting program then yields:
\begin{tabbing}
MMMMMMMMMMMMMMMMMMMMMMMMMM \= M \= \kill
$\pi_{sum1}, \pi_{gts }, \pi_{sum2}$
    \> $=$ \> $0$ \\
$\pi_{ys1 }, \pi_{ys2 }$
    \> $=$ \> $1$ \\
$c_{gts}, c_{ys1}, c_{ys2}$           
    \> $=$ \> $0$
\end{tabbing}
Note that although nodes $sum1$ and $sum2$ have equal $\pi$ values, they are not fused. Conversely, if two nodes have different $\pi$ values then they are never fused. The corresponding value of the objective function is: $25 \cdot x_{sum1, gts} + 1 \cdot x_{sum1,sum2} + 25 \cdot x_{sum1, ys2} + 25 \cdot x_{gts, ys1} + 1 \cdot x_{sum2, ys1} + 25 \cdot x_{ys1, ys2} = 102$.

% -----------------------------------------------------------------------------
\subsection{Closest Points}
The closest points benchmark is a divide-and-conquer algorithm that, given an array of points, finds the closest pair of points.
A midpoint is found, and the points are filtered into those above the midpoint, and those below.
The closest points in the filtered arrays are then recursively computed, and the two results merged.
As the filters are operated on recursively, there are no opportunities for fusing the filters, making our clustering the same as Megiddo's.
Our clustering is able to generate both filtered arrays in a single loop, unlike stream fusion, which requires a separate loop for each.

% -----------------------------------------------------------------------------
\subsection{QuadTree}
The QuadTree benchmark recursively computes a 2-dimensional space partitioning tree of an array of points.
At each step, the array of points is filtered into four boxes.
As with closest points, there are no opportunities for fusing the filters, and our clustering is the same as Megiddo's.
Our clustering produces all four filtered results in a single loop, whereas stream fusion requires four loops.

% -----------------------------------------------------------------------------
\subsection{QuickHull}
QuickHull is another program that is worth mentioning, as it highlights some fusion opportunities with filters.
The core of the QuickHull algorithm is shown below: given a line and an array of points, the points are filtered to those above the line, and the farthest point above the line is also found.
